---
title: "Two-way Hierarchical Analysis of Variance by using the restricted maximum likelihood(REML) "
author: "Yue Zhang and Paul Northrop"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Two-way Hierarchical Analysis of Varianceby using the restricted maximum likelihood(REML)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: climr.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The 'fit_reml' function is only applicable for the CMIP5 data containing the index, GCM and RCP; This function is also for fitting a random-effects ANOVA model and finding the standard deviations of estimators by using the restricted maximum likelihood (REML) estimation, following @Northrop_2014.

# Text to introduce the structure of the (CMIP5) data

According to the description in @Taylor_2012, the CMIP5 is the fifth phase of the Coupled Model Intercomparison Project for understanding of past and future climate changes. The main experiments are long term (century time scale) integration and near-term integration (10-30 yr). In addition, there are four CMIP5 scenarios that are RCP 2.6, RCP 4.5, RCP 6.0 and RCP 8.5. However, there are some basic considerations for users of CMIP5 data, such as unforced variability and climate drift and bias correction. 

The 'cmip5_temp1' data contains the mid 21st century global temperature projection data. At the same time, the 'cmip5_temp2' data contains late 21st century global temperature projection data. Both of them have 270 rows and 4 columns.Each row relates to a climate projection run from one of 38 different General Circulation Models (GCMs) under a particular Representative Concentration Pathway (RCP).Column 1 contains the anomaly of the mean global temperature between two different time period. For 'cmip5_temp1', it is the mean global temperature over the time period 2020-2049 relative to that over 1970-1999 and for 'cmip5_temp2', it is the mean global temperature over the time period 2020-2049 relative to that over 1970-1999 i.e. the latter subtracted from the former; Column 2 contains an abbreviation for the name of the climate modelling research group and the GCM; Column 3 contains the one of four RCPs; Column 4 is the simulation run number.


```{r setup}
library(climr)

# showing data that stored in the pacake (climr)
head(cmip5_temp1)
head(cmip5_temp2)
```

# Formula for the Two-way random-effected ANOVA model
Suppose $Y_{ijk}$; i = 1,...,$n_{GCM}$; j = 1,...,$n_{RCP}$; and k = 1,...,$K_{ij}$ be an index of change for GCM $i$, RCP $j$ and run $k$. For the CMIP5 data, $n_{GCM}$  = 38, $n_RCP$ = 4 and $K_{ij}$ varies from 0 to 13. Here is the statistical model:
$$Y_{ijk} = \mu + \alpha_i + \beta_j + \gamma_{ij} + \epsilon_{ijk}, $$
$$i = 1,...,n_{GCM}, j = 1,...,n_{RCP}, k = 1,...,K_{ij},$$
$$\alpha_i \sim N(0,\sigma_{GCM}^2), \beta_j \sim N(0,\sigma_{RCP}^2), \gamma_{ij} \sim N(0,\sigma_{GCM:RCP}^2), \epsilon_{ijk} \sim N(0,\sigma_{R}^2).$$

The $\mu$ is the overall mean change in the index over all GCMs and RCPs and is the fixed effect. And the  $\alpha_i$, $\beta_j$ and $\gamma_{ij}$ are the random effects which are normally distributed with mean 0 and corresponding variance. For example, the $\alpha_i$ interprets an adjustment to GCM $i$ with variance $\sigma_{GCM}^2$; With the same notation, $\beta_j$ interprets an adjustment to RCP $j$ with variance  $\sigma_{RCP}^2$ and the $\gamma_{ij}$ interprets interaction between the GCM $i$ and RCP $j$ that also indicates a RCP-specific additional adjustment for GCM $i$ with variance $\sigma_{GCM:RCP}^2$. Then $\epsilon_{ijk}$ is the error term that is also normally distributed with mean 0 and variance $\sigma_{R}^2$, meaning the residual variability between runs which also is the variability within the modeled system. All the random variables are independent in the assumption.

Due to the unbalance of data and the random variables in model, we would Use the restricted maximum likelihood (REML) estimation. The parameters are  $\mu, \sigma_{GCM}, \sigma_{RCP}, \sigma_{GCM:RCP}$ and $\sigma_{R}$ that are treated as $\theta$:
$$(\mu,\sigma_{GCM},\sigma_{RCP},\sigma_{GCM:RCP},\sigma_{R}) = \theta$$

Using the simple maximum likelihood maximise the likelihood function of $\theta$ (i.e. $L(\theta)$) with respect to $\theta$ but the REML would treat the fixed-effect variable ($\mu$) and the random-effect variables ($\sigma_{GCM}, \sigma_{RCP}, \sigma_{GCM:RCP}, \sigma_{R}$) separately and estimated the previous one firstly.

# Explain the purpose of the REML

The data would fit in a two-way random-effects ANOVA model
containing three factors which are GCM, RCP and the interaction between GCM
and RCP. Those factors are all considered to be normally distributed random
variables with mean zero and different standard deviation, like
\eqn{sigma_GCM,sigma_RCP,sigma_GCM_RCP}. In addition, the intercept
\eqn{\mu} of this model interprets the overall mean change in the index and
the error term \eqn{residual} are independent identically distributed
random variables with mean zero and standard deviation \eqn{sigma_run}. 
More information see @lme4.

This function uses the restricted maximum likelihood (REML) method
for estimating those estimators, their corresponding standard errors and their 
corresponding 95% confidence intervals.

```{r, fig.show='hold', fig.width = 3.45, fig.height = 3.45}
x <- fit_reml(cmip5_temp1)
y <- fit_reml(cmip5_temp2)
```

The return is an object (a list) of class 'climr_reml', 'cmip5'and 'climr' containing the result of fitted model using the REML method from \code{lme4::lmer}, named \code{lmer_object}. Besides, there are the estimates of standard deviations of random-effects factors and the residual, including their each standard error and 95% confidence interval shown by upper limits (i.e. 2.5%) and lower limits (i.e. 97.5%).

```{r, fig.show='hold', fig.width = 3.45, fig.height = 3.45}
# the call of 'fit_reml'
x$call
y$call

# the result of fitted model using the REML method from lme4::lmer
x$lmer_object
y$lmer_object

# the estimates of standard deviations of random-effects factors and the residual
x$ests
y$ests

# the standard errors of those estimates
x$ses
y$ses

# the 95% confidence interval for those estimates
x$ci
y$ci
```

## Difference formats of showing the results

There is a function named 'print.climr_reml' for print the results.

```{r, fig.show='hold', fig.width = 3.45, fig.height = 3.45}
print(x)
print(y)
```

There is a function named 'summary.climr_reml' for summary the results.

```{r, fig.show='hold', fig.width = 3.45, fig.height = 3.45}
summary(x)
summary(y)
```

There is a function named 'print.summary.climr_reml' for print the summary of results.

```{r, fig.show='hold', fig.width = 3.45, fig.height = 3.45}
print(summary(x))
print(summary(y))
```
# References

